---
title: "Childcare Cost Analysis"
subtitle: "Project 1 - Proposal"
author:
  - name: "**Team InsightSquad** - Mohammad Farmani, Narasimha Vardhan Rachaputi, Amaan"
  - name: "Devendran, Shreemithra, Gowtham"
format: html
editor: visual
execute:
  warning: false
  message: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

## Introduction

The objective of this proposal is to explore the complex dynamics between childcare costs, unemployment rates, and poverty levels across various locations in the United States. By utilizing two main datasets—the National Database of Childcare Prices (NDCP) and a comprehensive Counties Dataset—we aim to uncover patterns and correlations that can inform us about any hidden relationship between these variables. The NDCP dataset, spanning from 2008 to 2018, offers an unparalleled look at childcare prices across the county level, categorized by provider type and age groups, while the Counties Dataset provides essential geographical and demographic insights.

## Loading Dataset

```{r}
#| label: load-dataset
#| message: false
library(here)
child_cost <- read.csv(here('Data','childcare_costs.csv'))
child_cost|>glimpse()
```

## Dataset Information

### 1. [Childcare Costs Dataset](https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-05-09/childcare_costs.csv)

-   **Source:** The National Database of Childcare Prices (NDCP)
-   **Description:** This dataset is the most comprehensive federal source of childcare prices at the county level. It provides childcare price data by childcare provider type, age of children, and county characteristics. The data spans from 2008 to 2018, offering estimates of childcare prices at the county level for different age groups and care settings, including home-based and center-based providers.

### 2. [Counties Dataset](https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-05-09/counties.csv)

-   **Description:** This dataset contains geographical information for each county, and it will be used for spatial plotting.

| Variable       | Description                                                                                                                |
|-------------------|-----------------------------------------------------|
| unr_16         | Unemployment rate of the population aged 16 years old or older.                                                            |
| funr_16        | Unemployment rate of the female population aged 16 years old or older.                                                     |
| munr_16        | Unemployment rate of the male population aged 16 years old or older.                                                       |
| pr_f           | Poverty rate for families.                                                                                                 |
| pr_p           | Poverty rate for individuals.                                                                                              |
| mc_infant      | Aggregated weekly, full-time median price charged for Center-based Care for infants (i.e. aged 0 through 23 months).       |
| mc_toddler     | Aggregated weekly, full-time median price charged for Center-based Care for toddlers (i.e. aged 24 through 35 months).     |
| mc_preschool   | Aggregated weekly, full-time median price charged for Center-based Care for preschoolers (i.e. aged 36 through 54 months). |
| mfcc_infant    | Aggregated weekly, full-time median price charged for Family Childcare for infants (i.e. aged 0 through 23 months).        |
| mfcc_toddler   | Aggregated weekly, full-time median price charged for Family Childcare for toddlers (i.e. aged 24 through 35 months).      |
| mfcc_preschool | Aggregated weekly, full-time median price charged for Family Childcare for preschoolers (i.e. aged 36 through 54 months).  |

### Reason for Choosing this Dataset

The reason for choosing this dataset is rooted in its potential to offer comprehensive insights into the intersection of childcare prices, unemployment rates, and poverty levels across different locations in the United States. The dataset covers a significant timeframe (2008 to 2018) and includes variables such as unemployment rates for different demographics, poverty rates for families and individuals, and median prices for center-based and family childcare across various age groups.

## Questions

### Question 1:

#### Which locations have lower unemployment rates? Which gender experiences higher unemployment rates in the US? Is there a relationship between the unemployment rate and poverty?

### Question 2:

#### Which places have higher rates of family poverty? Is there a relationship between poverty and the cost of raising children?

## Analysis Plan

#### Introduction

This section outlines our methodology for investigating the relationships between unemployment rates, poverty levels, and childcare costs across the United States. By examining these variables, we aim to identify patterns that could have significant implications for social policy and economic support systems.

#### Approach

To systematically address our research questions, we will employ spatial analysis techniques alongside statistical correlation assessments. The analysis will be conducted in two main parts, corresponding to our central questions:

#### For Question 1

-   We will focus on the distribution of unemployment rates across genders and its relationship with poverty. The methodology includes creating spatial maps to visualize mean unemployment rates (overall, and by gender) across the US. We will then calculate the correlation between the overall unemployment rate and individual poverty levels at the county level, producing spatial maps to display these correlations.

#### For Question 2

-   The emphasis will be on understanding the dynamics between family poverty rates and the financial burden of childcare. This part involves generating spatial maps to illustrate the average rate of family poverty across the US. Subsequently, we will assess the correlation between family poverty rates and the cumulative childcare costs (encompassing both Family Childcare and Center-based Care) for all age groups at the county level, with spatial maps visualizing these correlations.

#### Discussion

The proposed analysis plan is designed to dissect the intricate web of economic factors influencing family well-being and child-rearing costs. By delving into the spatial and statistical relationships between unemployment, poverty, and childcare expenses, we anticipate uncovering regional disparities and potential areas for policy intervention. The findings from this analysis could provide a robust evidence base for developing targeted strategies to alleviate the economic pressures on families, particularly in regions where the burden of childcare costs significantly intersects with high unemployment and poverty rates. Through this investigation, we aim to contribute to the broader dialogue on enhancing economic security and access to quality childcare, thereby supporting the well-being of families across the United States.

## Timeline

-   Revise and complete the proposal before Monday, Feb 19th.

-   Complete the coding part and generating the plots for both questions before Monday, Feb 26th.

-   Complete the presentation before Monday, March 4th.
