---
title: "Childcare Cost Analysis"
subtitle: "Project 1 - Proposal"
author: 
  - name: "Team InsightSquad - Mohammad Farmani, Narasimha Vardhan Rachaputi, Amaan, Devendran, Shreemithra, Gowtham"
format: html
editor: visual
execute: 
  warning: false
  message: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false
library(here)
child_cost <- read.csv(here('Data','childcare_costs.csv'))
child_cost|>glimpse()
```

## About the Dataset

The National Database of Childcare Prices (NDCP) is the most comprehensive federal source of childcare prices at the county level. The database offers childcare price data by childcare provider type, age of children, and county characteristics. Data are available from 2008 to 2018. It provides estimates of childcare prices at the county level by children’s age groups and care settings (home-based or center-based providers).

The dataset comprises various variables, making it suitable for diverse analyses. Although it lacks categorical variables, we can transform some of the variables into categorical format using bins. Moreover, this dataset offers valuable insights into poverty, race, population, and the unemployment rate across the US, enhancing our understanding of these issues. 


We will use some of the provided data to do our analysis, not the whole available data in the files, and the main variables that will be used and their description are provided in the following table. It should be noted there is another CSV file containing the gepgraphical information of each counties, and we will use it for spatial plotting.

|                |                                                                                                                            |
|----------------|----------------------------------------------------------------------------------------------------------------------------|
| Variable       | Description                                                                                                                |
| unr_16         | Unemployment rate of the population aged 16 years old or older.                                                            |
| funr_16        | Unemployment rate of the female population aged 16 years old or older.                                                     |
| munr_16        | Unemployment rate of the male population aged 16 years old or older.                                                       |
| pr_f           | Poverty rate for families.                                                                                                 |
| pr_p           | Poverty rate for individuals.                                                                                              |
| mc_infant      | Aggregated weekly, full-time median price charged for Center-based Care for infants (i.e. aged 0 through 23 months).       |
| mc_toddler     | Aggregated weekly, full-time median price charged for Center-based Care for toddlers (i.e. aged 24 through 35 months).     |
| mc_preschool   | Aggregated weekly, full-time median price charged for Center-based Care for preschoolers (i.e. aged 36 through 54 months). |
| mfcc_infant    | Aggregated weekly, full-time median price charged for Family Childcare for infants (i.e. aged 0 through 23 months).        |
| mfcc_toddler   | Aggregated weekly, full-time median price charged for Family Childcare for toddlers (i.e. aged 24 through 35 months).      |
| mfcc_preschool | Aggregated weekly, full-time median price charged for Family Childcare for preschoolers (i.e. aged 36 through 54 months).  |

## Questions

#### Question 1: Which locations have lower unemployment rates? Which gender experiences higher unemployment rates in the US? Is there a relationship between the unemployment rate and poverty?

#### Question 2: Which places have higher rates of family poverty? Is there a relationship between poverty and the cost of raising children?

## Analysis plan

-   For the first question, we will create spatial maps displaying the mean overall unemployment rate as well as the mean unemployment rates for females and males across the US. Subsequently, for each county, we will calculate the correlation between the overall unemployment rate and individual poverty levels and a spatial map of these correlations will be plotted.

-   For the second question, We begin by mapping the average family poverty rates across the US using a Spatial map. Subsequently, compute the correlation between family poverty and total child-rearing costs, including Family Childcare and Center-based Care, for each county. Then, we introduce a scatter plot to visualize this relationship, with family poverty rates on the x-axis and child-rearing costs on the y-axis, plotting each county as a point. Then analyze the scatter plot to identify trends and consider adding a trend line for clarity. Optionally, overlay the scatter plot onto the spatial map to juxtapose the relationship with the distribution of family poverty rates across counties. This integrated approach provides a comprehensive understanding of the relationship between family poverty and child-rearing costs while leveraging spatial visualization techniques for enhanced insights.
