---
title: "Childcare Cost Analysis"
subtitle: "Project 1 - Proposal"
author:
  - name: "**Team InsightSquad** - Mohammad Farmani, Narasimha Vardhan Rachaputi, Amaan"
  - name: "Devendran, Shreemithra, Gowtham"
format: html
editor: visual
execute:
  warning: false
  message: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false
library(here)
child_cost <- read.csv(here('Data','childcare_costs.csv'))
child_cost|>glimpse()
```

# Dataset Information

### 1. [Childcare Costs Dataset](https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-05-09/childcare_costs.csv)

-   **Source:** The National Database of Childcare Prices (NDCP)
-   **Description:** This dataset is the most comprehensive federal source of childcare prices at the county level. It provides childcare price data by childcare provider type, age of children, and county characteristics. The data spans from 2008 to 2018, offering estimates of childcare prices at the county level for different age groups and care settings, including home-based and center-based providers.

### 2. [Counties Dataset](https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-05-09/counties.csv)

-   **Description:** This dataset contains geographical information for each county, and it will be used for spatial plotting.

| Variable       | Description                                                                                                                |
|-------------------|-----------------------------------------------------|
| unr_16         | Unemployment rate of the population aged 16 years old or older.                                                            |
| funr_16        | Unemployment rate of the female population aged 16 years old or older.                                                     |
| munr_16        | Unemployment rate of the male population aged 16 years old or older.                                                       |
| pr_f           | Poverty rate for families.                                                                                                 |
| pr_p           | Poverty rate for individuals.                                                                                              |
| mc_infant      | Aggregated weekly, full-time median price charged for Center-based Care for infants (i.e. aged 0 through 23 months).       |
| mc_toddler     | Aggregated weekly, full-time median price charged for Center-based Care for toddlers (i.e. aged 24 through 35 months).     |
| mc_preschool   | Aggregated weekly, full-time median price charged for Center-based Care for preschoolers (i.e. aged 36 through 54 months). |
| mfcc_infant    | Aggregated weekly, full-time median price charged for Family Childcare for infants (i.e. aged 0 through 23 months).        |
| mfcc_toddler   | Aggregated weekly, full-time median price charged for Family Childcare for toddlers (i.e. aged 24 through 35 months).      |
| mfcc_preschool | Aggregated weekly, full-time median price charged for Family Childcare for preschoolers (i.e. aged 36 through 54 months).  |

# Reason for Choosing this Dataset

The reason for choosing this dataset is rooted in its potential to offer comprehensive insights into the intersection of childcare prices, unemployment rates, and poverty levels across different locations in the United States. The dataset covers a significant timeframe (2008 to 2018) and includes variables such as unemployment rates for different demographics, poverty rates for families and individuals, and median prices for center-based and family childcare across various age groups.

## Questions

### Question 1:

#### Which locations have lower unemployment rates? Which gender experiences higher unemployment rates in the US? Is there a relationship between the unemployment rate and poverty?

### Question 2:

#### Which places have higher rates of family poverty? Is there a relationship between poverty and the cost of raising children?

## Analysis Plan

-   For the first question, we will create spatial maps displaying the mean overall unemployment rate as well as the mean unemployment rates for females and males across the US. Subsequently, for each county, we will calculate the correlation between the overall unemployment rate and individual poverty levels, and a spatial map of these correlations will be plotted.

-   For the second question, we will generate spatial maps showing the average family poverty across the US. Then, for each county, we will calculate the correlation between family poverty and the total costs for children of all ages including Family Childcare and Center-based Care. A spatial map illustrating these correlations will be plotted.

### Timeline

-   Revise and complete the proposal before Monday, Feb 19th.

-   Complete the coding part and generating the plots for both questions before Monday, Feb 26th.

-   Complete the presentation before Monday, March 4th.
